---
title: "FinTech Challenge"
author: "Mohamed Khalifa"
output:
  pdf_document: default
  html_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

setwd("C:/Users/hb92148/Documents/R_projects/FinTech")


```


###Code Structure:
#### 1. Importing data
#### 2. Data Manipulation
#### 3. Feature Extraction
#### 4. Graphs

<br>




##1. Importing data

In this section, the goal is loading the dataset given for this exercise.


```{r}

# Loading libraries

library(tidyverse)
library(lubridate)


# Importing data in proper view 
df <- read.csv("challenge_data.csv", as.is = T, check.names = F, na.strings = "NULL")
colnames(df) <- gsub(" ", "_", colnames(df)) 
colnames(df) <- gsub("-", "_", colnames(df))
colnames(df) <- tolower(colnames(df))

str(df)

```


### Order data ascending by offer_creation_date column

```{r}

df <- df[order(df$offer_creation_date), ]

```


## Missing values

```{r}

data.frame(NA_Count = sort(colSums(is.na(df)) , decreasing = T ))

```



##2. Data Manipulation 


### Changing ID columns 

In this section, we will change id - character- columns to more user fiendly pattern for easier processing.
The columns are in random characters format that implies no useful structure beind.

<b>The characters columns to be changed:<b> <br>
- Offer ID <br>
- Payment Job ID <br>
- Transfer Job ID <br>
- User ID 

```{r}

## Common steps in every column reformating:
# 1. Convert the column class from character to factor.
# 2. Create sequence with the unique values length in every column.
# 3. Concatenating the intial of the column name with the sequence already created.
# 4. Returning the column class into the intial character format.



# offer_id column

df$offer_id <- as.factor(df$offer_id)
levels(df$offer_id) <- paste0("o", "_", seq(length(unique(df$offer_id))))
df$offer_id <- as.character(df$offer_id)


# payment_job_id column

df$payment_job_id <- as.factor(df$payment_job_id)
levels(df$payment_job_id) <- paste0("p", "_", seq(length(unique(df$payment_job_id))))
df$payment_job_id <- as.character(df$payment_job_id)


# transfer_job_id column

df$transfer_job_id <- as.factor(df$transfer_job_id)
levels(df$transfer_job_id) <- paste0("t", "_", seq(length(unique(df$transfer_job_id))))
df$transfer_job_id <- as.character(df$transfer_job_id)


# user_id column

df$user_id <- as.factor(df$user_id)
levels(df$user_id) <- paste0("u", "_", seq(length(unique(df$user_id))))
df$user_id <- as.character(df$user_id)

```


### Detecting unique values in ID columns

```{r}

length(unique(df$offer_id))
length(unique(df$payment_job_id))
length(unique(df$transfer_job_id))
length(unique(df$user_id))

```


### Converting Date columns 

In this section, we will change all Date columns from character format to POXISct format for easier processing.

<b>The Date columns to be changed:<b> <br>
01- offer_creation_date <br>
02- offer_expiration_date <br>
03- payment_job_creation_date <br>
04- payment_job_closure_date <br>
05- voucher_buying_date <br>
06- user_registration_date <br>
07- user_first_login <br>
08- user_last_login <br>
09- user_last_offer_date <br>
10- user_first_voucher_buying_date <br>
11- user_last_voucher_buying_date <br>

```{r}

date_cols <- c()  # comment_1

for(i in seq(ncol(df))){
  if ( (class(df[, i]) == "character") & (any(grepl("\\:", df[, i])) == T)){   # comment_2
    date_cols <- c(date_cols, colnames(df[i]))
  }
}

df[date_cols] <- lapply(df[date_cols], as.POSIXct, format= "%m/%d/%Y %H:%M")



# comment_1: Create vector for all the Date columns
# comment_2: The grepl function was used to ensure that the column is of Date type not a usual character column

```


### Converting digital columns 

In this section, we will change all digits columns to numeric format for easier processing.

The Date columns to be changed: <br>
01- offer_amount <br>
02- nominal_value <br>
03- upside_amount <br>
04- user_total_logins <br>
05- user_total_emails <br>
06- user_total_offers 

```{r}


for( i in colnames(df)){
  if( !(i %in% c(date_cols, "offer_id", "payment_job_id", "transfer_job_id", "user_id"))){   # comment_1
    df[ , i] <- as.numeric(df[ , i] )
  }
} 

# comment_1: Date columns and ID -character- columns have been excluded from the loop.

str(df)

```


##3. Feature Extraction

###3.1 Create offer_duration column
It's the days difference between offer_expiration_date & offer_creation_date column  column

```{r}

df$offer_duration <- round(difftime(df$offer_expiration_date, df$offer_creation_date, units = "days"), 1)  #comment_1
df$offer_duration[which(df$offer_duration == 0.1)] <- 0.6  #comment_2

df$offer_duration <- as.factor(df$offer_duration)



#comment_1: difference is rounded for easier factorization
#comment_2: a unique 0.1 factor has been considered as 0.6



summary(as.factor(df$offer_duration))

```


###3.2 Create diff_bet_login__voucher_bought column

```{r}
# It's the days difference between user_first_voucher_buying_date & user_first_login

df$diff_bet_login__voucher_bought <- round(as.numeric
                                           (difftime(df$user_first_voucher_buying_date, df$user_first_login, units = "days")))

```


###3.3 Create offers_per_mail column
It's the division of user_total_offers over user_total_e_mails

```{r}

df$offers_per_mail <- round(df$user_total_offers / df$user_total_e_mails , 1)

```


###3.4 Create offer_weekday column
It's the weekdays of the offer_creation_date column

```{r}

df$offer_weekday <- wday(df$offer_creation_date, label=TRUE)

```


###3.5 Create register_weekday column
It's the weekdays of the user_registration_date column

```{r}

df$register_weekday <- wday(df$user_registration_date, label=TRUE)

```


###3.6 Create offer_taken column
It's a binary factor column for a completely transfered job. <br>
It's validated by the presence of a transfer_job_id. <br>
Factor (1) means that the offer has been acquired by the user. Factor (0) denotes the opposite

```{r}

df$offer_taken <- lapply( df$transfer_job_id, function(x) ifelse((is.na(x) == F), 1, 0))
df$offer_taken <- as.factor(unlist(df$offer_taken))

```


###3.7 Create login_weeks column
It's the difference in weeks between user_last_login & user_first_login

```{r}

df$login_weeks <- round(difftime(df$user_last_login, df$user_first_login, units = "weeks"))

```


###3.8 Create login_frequency column
It's the division of user_total_logins over login_weeks

```{r}

for ( i in seq(nrow(df))){
  if (df$login_weeks[i] != 0 ){
    df$login_frequency[i] <- round(df$user_total_logins[i] / as.numeric(df$login_weeks[i]), 1)
  }
  else{
    df$login_frequency[i] <- 0
  }
}

```


###3.9 Create diff_bet_1st__last_vouchers column

```{r}
# It's the days difference between user_last_voucher_buying_date & user_first_voucher_buying_date

df$diff_bet_1st__last_voucher <- round(
  as.numeric(difftime(df$user_last_voucher_buying_date, df$user_first_voucher_buying_date, units = "days")))

```


###3.10 Create diff_bet_voucher_bought__used column

```{r}
# It's the days difference between voucher_buying_date & user_first_voucher_buying_date

df$diff_bet_voucher_bought__used <- round(
  as.numeric(difftime(df$voucher_buying_date, df$user_first_voucher_buying_date, units = "days")))

```


###3.11 Create offer_percent column
It's the division of upside_amount over nominal_value. <br>
The column has been further processed to 10 quantiles from 10% to 100%

```{r}

df$offer_percent <- round(df$upside_amount / df$nominal_value, 2)
df$offer_percent <- cut(df$offer_percent, breaks = 10, labels= 1 : 10)

levels(df$offer_percent) <- seq(10, 100, 10)

```


###3.12 Create transfer_hour column
It's extracted from the voucher_buying_date column

```{r}

for(i in seq(nrow(df))){
  if(is.na(df$voucher_buying_date[i]) != T){
    df$transfer_hour[i] <- hour(df$voucher_buying_date[i])
    df$transfer_hour[i] <- as.numeric(df$transfer_hour[i])
  }
  else{
    df$transfer_hour[i] <- NA
  }
}

```


###3.13 Create payment_weekday column
It's the weekdays of the payment_job_closure_date column

```{r}

df$payment_weekday <- wday(df$payment_job_closure_date, label=TRUE)

```



### Reordering the columns

```{r}
col_names <- c("offer_id", "offer_creation_date", "offer_weekday", "offer_amount", "offer_expiration_date", 
               "offer_duration", "payment_job_id", "payment_job_creation_date", "payment_job_closure_date", 
               "payment_weekday", "transfer_job_id", "voucher_buying_date", "transfer_hour", 
               "diff_bet_voucher_bought__used", "nominal_value", "upside_amount", "offer_percent", 
               "user_id", "user_registration_date", "register_weekday", "user_first_login", "user_last_login", 
               "login_weeks", "user_total_logins", "login_frequency", "user_total_e_mails", 
               "user_last_offer_date", "user_total_offers", "offers_per_mail", "diff_bet_login__voucher_bought", 
               "user_first_voucher_buying_date", "user_last_voucher_buying_date", "diff_bet_1st__last_voucher",
               "offer_taken")

df <- df[, col_names]

```



##4. Graphs


###4.1 Number of users registered per month

```{r}

# Chronological months order
users_reg_months <- data.frame(df %>%
                                 group_by(months = floor_date(user_registration_date, "month"))  %>%
                                 summarise(users_per_month = length(unique(user_id))) %>% 
                                 mutate(months_names = c("Dec '17", "Jan", "Feb", "Mar", "Apr", "May", 
                                                         "Jun", "Jul", "Aug", "Sep", "Oct")))


ggplot( data = users_reg_months[which(users_reg_months$months_names != "Oct"), ],  # comment_1 
        aes(x= fct_inorder(months_names) , y = users_per_month)
        ) + 
  geom_bar(stat = "identity", fill = "steelblue", width = 0.7) + 
  labs(title = "Users Registration Count per Month", 
       y = "Users") +
  theme(plot.title = element_text(size = 15, hjust = 0.5),
        axis.title.x = element_blank(), 
        text = element_text(size = 15)) +
  geom_text(aes(label = users_per_month), 
            vjust = 1.6, color = "white", size = 3.5)



# Descending users count per month order 
users_reg_months <- data.frame(df %>%
                                 group_by(months = floor_date(user_registration_date, "month"))  %>%
                                 summarise(users_per_month = length(unique(user_id))) %>% 
                                 mutate(months_names = c("Dec", "Jan", "Feb", "Mar", "Apr", 
                                                         "May", "Jun", "Jul", "Aug", "Sep", "Oct"))) %>% 
                                arrange(desc(users_per_month))


ggplot( data = users_reg_months[which(users_reg_months$months_names != "Oct"), ], 
        aes(x= reorder(months_names, users_per_month ), y = users_per_month)
        ) + 
  geom_point(size = 4, shape = 19, color = "tomato") +
  geom_segment(aes(x = months_names, 
                   xend = months_names, 
                   y = 0, 
                   yend = users_per_month), 
               color = "steelblue", size = 1) + 
  labs(title = "Users Registration Count per Month",
       y = "Users") + 
  theme(plot.title = element_text(size = 15, hjust = 0.5),
        axis.title.y = element_blank(), 
        text = element_text(size = 15)) + 
  coord_flip()



# Descending users count per first 9 days per month order
users_reg_months_9days <- df %>% filter(mday(user_registration_date) < 8.5) %>% 
  group_by(months_names = floor_date(user_registration_date, "month"))  %>% 
  summarise(users_per_month = length(unique(user_id))) %>% 
  mutate(months_names = c("Dec'17", "Jan", "Feb", "Mar", "Apr", "May", "Jun", "Jul", "Aug", "Sep", "Oct")) %>%
  arrange(desc(users_per_month))

bars_colour <- c("steelblue", "steelblue", "steelblue", "steelblue", "steelblue", 
                 "steelblue", "steelblue", "steelblue", "tomato", "steelblue", "steelblue")
ggplot( data = users_reg_months_9days, 
        aes(x= reorder(months_names, users_per_month ), y = users_per_month)
        ) + 
  geom_bar(stat = "identity", width = 0.7, fill = bars_colour) + 
  labs(title = paste("Users Registration Count per Month \n (First 9 Days each Month)"),
       y = "Users") + 
  geom_text(aes(label = users_per_month), hjust = -0.5, size = 4.5) +
  theme(plot.title = element_text(size = 15, hjust = 0.5),
        axis.title.y=element_blank(), 
        text = element_text(size = 15)) + 
  coord_flip()



# Mean registrations count for months prior to August (for presentation)
users_reg_months %>% filter(!months_names %in% c("Aug", "Sep", "Oct")) %>% summarise(mean(users_per_month))


# comment_1: October month will be removed because it's not complete

```


###4.2 Users offer acceptance per weekday

```{r}

offers_taken_per_day <- df %>% group_by(offer_weekday) %>% filter(offer_taken == 1) %>%
  summarise(total = n()) %>% arrange(desc(total))
offers_taken_per_day <- data.frame(offers_taken_per_day)

total_offers_per_day <- df %>% group_by(offer_weekday) %>% summarise(total = n()) %>% arrange(desc(total))
total_offers_per_day <- data.frame(total_offers_per_day); total_offers_per_day


offers_taken_percent_per_day <- data.frame(prop.table(table(df[ , c("offer_weekday", "offer_taken")]), 1))
offers_taken_percent_per_day$Freq <- round(offers_taken_percent_per_day$Freq, 3)
offers_taken_percent_per_day <- offers_taken_percent_per_day[order(offers_taken_percent_per_day$Freq, decreasing = T), ]
levels(offers_taken_percent_per_day$offer_taken) <- c("Rejected percent", "Accepted percent")

offers_taken_percent_per_day <- offers_taken_percent_per_day %>% mutate(ordering = row_number())


# total_offers per day barplot
bars_colour <- c("steelblue", "steelblue", "steelblue", "steelblue", "steelblue", "steelblue", "tomato")

ggplot(data = total_offers_per_day, 
       aes(x = reorder(offer_weekday, -total), y = total)
       ) +
  geom_bar(aes( fill = "Sun"), stat = "identity", width = 0.5, color = "black", fill = bars_colour) +
  geom_text(aes(label = total) , vjust = -0.3, size = 3.9) + 
  labs(title = "Total offers per day",
       x = "Weekdays") + 
  scale_y_continuous (expand = c(0, 0), limits = c(0, 27000)) +
  theme( plot.title = element_text(size = 16, hjust = 0.5),
         axis.title.x = element_text(size = 11, face = "bold"),
         axis.text.x = element_text(size = 11),
         axis.title.y = element_blank(),
         axis.text.y = element_text(size = 11))


# offers_taken percent per day barplot
bars_colour <- c("steelblue", "steelblue", "steelblue", "steelblue", "steelblue", "steelblue", "tomato",
                 "tomato", "steelblue", "steelblue", "steelblue", "steelblue", "steelblue", "steelblue")

ggplot(data = offers_taken_percent_per_day, 
       aes(x = ordering, y = Freq)
       ) + 
  facet_grid(~ offer_taken, scales = "free") +
  scale_x_continuous( breaks = offers_taken_percent_per_day$ordering, 
                      labels = offers_taken_percent_per_day$offer_weekday) +
  geom_bar(stat = "identity", width = 0.5, color = "black", fill = bars_colour) +
  geom_text(aes(label = Freq) , 
            vjust = -0.3, size = 3.5) + 
  labs(title = "Offers acceptance percentage",
       x = "Weekdays") + 
  scale_y_continuous (expand = c(0, 0), limits = c(0, 1.05)) +
  theme(plot.title = element_text(size = 16, hjust = 0.5),
        axis.title.x = element_text(size=11, face="bold"),
        axis.text.x = element_text(size = 10.5),
        axis.title.y = element_blank(),
        axis.text.y = element_text(size = 10.5), 
        strip.text.x = element_text(size = 11.5) )

```


###4.3 Offer duration effect on offers taken

```{r}

prop.table(table(df[ , c("offer_duration", "offer_taken")]), 1)
# The offer duration seems that it doesn't have huge effect on offers taken

```


###4.4 Register weekday effect on offers taken

```{r}

offers_taken_percent_per_reg_wday <- data.frame(prop.table(table(df[ , c("register_weekday", "offer_taken")]), 1))
offers_taken_percent_per_reg_wday$Freq <- round(offers_taken_percent_per_reg_wday$Freq, 3)
offers_taken_percent_per_reg_wday <- offers_taken_percent_per_reg_wday[order(offers_taken_percent_per_reg_wday$Freq, decreasing = T), ]

offers_taken_percent_per_reg_wday %>% filter(offer_taken == 1) 

# The registered weekday seems that it doesn't have that effect on offers taken
```


###4.5 Total emails per total logins effect on offers taken
The median of total_emails & total_logins for offers taken has been plotted for optimization

```{r}

median_total_logins <- as.numeric(df %>% filter(offer_taken == 1 & is.na(user_total_e_mails) != T) %>% 
                                    summarise(median(user_total_logins)))
median_total_emails <- as.numeric(df %>% filter(offer_taken == 1 & is.na(user_total_e_mails) != T) %>% 
                                    summarise(median(user_total_e_mails)))


ggplot(data = df,
       aes(x = user_total_logins, y = df$`user_total_e_mails`)
       ) + 
  geom_point() +
  facet_grid(~ offer_taken) +
  geom_point(aes(x = median_total_logins, y = median_total_emails), 
             colour="red", size = 3) +
  labs(title = "User Total logins vs. Total e-mails",
       x = "Total logins",
       y = "Total e-mails") +
  theme(plot.title = element_text(size = 15, hjust = 0.5),
        text = element_text(size = 15))

```


###4.6 Users most preferred transfer hour among weekdays

```{r}

df$payment_weekday <- factor(df$payment_weekday, levels = c("Mon", "Tue", "Wed", "Thu", "Fri", "Sat", "Sun" ))

p <- ggplot(data = df[!is.na(df$transfer_hour), ], 
            aes(x = transfer_hour, fill = payment_weekday)
            ) +
  geom_density(alpha = 0.4) +
  scale_fill_brewer(palette="Dark2") +
  facet_grid(payment_weekday ~ .) +
  theme(plot.title = element_text(size = 16, hjust = 0.5),
        axis.title.x = element_text(size=11, face="bold"),
        axis.text.x = element_text(size = 10.5),
        axis.title.y = element_blank(),
        axis.text.y = element_text(size = 10.5), 
        strip.text.x = element_text(size = 11.5) )

str(ggplot_build(p)$data[[1]])
to_fill <- data_frame(
  x = ggplot_build(p)$data[[1]]$x,
  y = ggplot_build(p)$data[[1]]$y,
  payment_weekday = factor(ggplot_build(p)$data[[1]]$PANEL, levels = c(1,2,3,4,5,6,7), 
                           labels = c("Mon", "Tue", "Wed", "Thu", "Fri", "Sat", "Sun" )))

p + geom_area(data = to_fill[to_fill$y >= 0.065, ], 
              aes(x = x, y = y), fill = "black", alpha = 0.7)


```


###4.7 Day time effect with offers_amount on offer taken
Transfer hour column was converted to 3 period (morning, afternoon, and night) to study the effect
of the offer amount on transfers behavior.

```{r}

for(i in seq(nrow(df))){
  if(is.na(df$transfer_hour[i] == T) ){
    df$offer_daytime[i] <- NA
  }
  else if( df$transfer_hour[i] > 5 & df$transfer_hour[i] <= 11){
    df$offer_daytime[i] <- "morning"
  }
  else if (df$transfer_hour[i] > 11 & df$transfer_hour[i] <= 19){
    df$offer_daytime[i] <- "afternoon"
  }
  else{
    df$offer_daytime[i] <- "night"
  }
}


ggplot(data = df[ df$offer_amount < 5000 & df$offer_taken == 1, ],
       aes(x = offer_amount, fill = offer_daytime)
       ) +
  geom_density(alpha = 0.4) +
  facet_grid(offer_daytime ~ .) 


df$offer_daytime <- NULL


# The transfer day time doesn't seem to have an effect on offers taken
```


###4.8 Duration between users first login and buying first voucher effect on offers taken

```{r}

ggplot( data = df[!is.na(df$diff_bet_login__voucher_bought), ], 
        aes(x = diff_bet_login__voucher_bought, fill = fct_rev(offer_taken))) +
  geom_density() + 
  labs(title = "Duration bet. Users 1st login & 1st Voucher",
       x = "Days") +
  theme(legend.position = c(0.9, 0.67)) +
  scale_fill_manual(values=c("green","tomato")) +
  theme(plot.title = element_text(size = 13, hjust = 0.5),
        axis.title.y=element_blank(), 
        text = element_text(size = 15)) 

```




